---
title: "Day One - Assessment"
format:
  html:
    df-print: paged
    embed-resources: true
    code-fold: true
bibliography: references.bib
csl: vancouver.csl
execute: 
  warning: false
editor: visual
---

# Visualization

::: {.callout-note appearance="minimal"}
## Learning Objectives

At the end of the assessment, participants should be able to:

-   Use the native pipe (`|>`) and the `{ggplot2}` plus-sign (`+`) correctly

-   Create general plots using `{ggplot2}` package and corresponding grammar; in particular:

    -   select/use tidy-data for a plot

    -   understand the meaning of "eesthetics" (`aes`)

    -   know the use and the basic geometries (`geom_*`)

    -   ability to facet a plot (`facet`)

    -   be able to modify the basic plot style (`theme` and `labs`)

    -   be aware that axes can be modified (e.g., `coord_flip`) and geometries can be personalized (e.g., \`position = "dodge"\`)
:::

# Preamble

::: callout-important
## Instructions

This section is for context only. Nothing has to be done in this section by the participants.
:::

::: panel-tabset
## Instructions

## RStudio cloud

## Attribution

## Copyright & License
:::

# Environment preparation

::: callout-important
## Instructions

In this section, participants should attach to the R session the following packages using the \`library\` function:

-   `tidyverse`

-   `here`

-   `rio`
:::

```{r, }
#| label: setup
library(tidyverse)
library(here)
library(rio)

linelist <- here("data-raw/Copenhagen_clean.xlsx") |> 
  import() |> 
  mutate(across(where(is.character), fct))
```

# Data preparation

::: callout-important
## Instructions

This section is for context only. Nothing has to be done in this section by the participants.
:::

::: panel-tabset
## Context

### The Alert

On November 14th 2006 the director of a high school in Greater Copenhagen, Denmark, contacted the regional public health authorities to inform them about an outbreak of diarrhoea and vomiting among participants from a school dinner party held on the 11th of November 2006. Almost all students and teachers of the school (750 people) attended the party.

The first people fell ill the same night and by 14 November, the school had received reports of diarrhoeal illness from around 200 - 300 students and teachers, many of whom also reported vomiting.

### Your mission

Your group has been tasked with investigating this outbreak; you have just received the information above. Before you spring into action, sit together and make a plan. Think about the ten steps of an outbreak investigation and how they apply in this setting. What practical issues might occur?

The particular focus of this session should be on steps 1-4 of the ten steps of outbreak investigations, i.e. the steps you need to take *before* you sit down and analyse the data.

## Data import

The epidemiologists in the outbreak team decided to perform a retrospective cohort study in order to identify the food item that was the vehicle of the outbreak. The cohort was defined as students and teachers who had attended the party at the high school on 11th of November 2006.

A questionnaire was designed to conduct a survey on food consumption and on presentation of the illness. Information about the survey and a link to the questionnaire was circulated to students and teachers via the school's intranet with the request that everyone who attended the school party on 11th of November 2006 should fill in the questionnaire.

Practically all students and teachers check the intranet on a daily basis, because it is the school's main communication channel for information about courses, homework assignments, cancellation of lessons etc. The school's intranet was accessible for ill students or teachers from home so that everyone in the cohort could potentially participate and the response rate could be maximised. Additionally, the information about the investigation was also displayed on the screen in the main hall of the school.

### Exploring the data

Before you can begin analysing the data, you will need to become familiar with its contents and check for any errors, or values that don't make sense.

::: callout-warning
## Preprocessing

For this assessment all the data preprocessing is already done. Participants will find the data already in their environment named as `linelist`.
:::

In the following:

-   The variables are grouped by type (e.g. character or numeric)

-   n_missing shows the number of observations with missing values for each variable

-   complete_rate shows the proportion of observations that are not missing

-   `min` and `max` for character variables refer to the number of characters per string

-   `p0` and `p100` for numeric variables refer to minimum and maximum values, respectively.

-   For more details, see the help file by typing `?skimr::skim` in the console

```{r}
#| label: data-skim
skimr::skim(linelist)

```

## Case definition

**Variables of interest:**

Your exposure of interest is the school dinner party held on 11 November 2006 at 18:00. You may have noticed while skimming the data, that there is a binary variable called `meal`. This variable indicates whether people attended the school dinner party and ate a meal there, or not.

Other variables that will be helpful to include in your case definition are `onset_datetime` (hint: check that case onset date/time is **after** exposure) and symptom variables (hint: not everyone on the linelist fell ill). The symptoms included in the data set are:

-   `abdo` (abdominal pain)
-   `diarrhoea`
-   `bloody` (bloody diarrhoea)
-   `nausea`
-   `vomiting`
-   `fever`
-   `headache`
-   `jointpain`

### Defining a case

A case was defined as a person who:

-   attended the school dinner on 11 November 2006 (i.e. is on the linelist)
-   ate a meal at the school dinner (i.e. was exposed)
-   fell ill after the start of the meal
-   fell ill no later than two days after the school dinner
-   suffered from diarrhoea with or without blood, or vomiting

Non cases were defined as people who:

-   attended the school dinner on 11 November 2006 (i.e. are on the linelist)
-   ate a meal at the school dinner (i.e. were exposed)
-   did not fall ill within the time period of interest
-   did not develop diarrhoea (with or without blood) or vomiting

For the sake of the analysis, we excluded any people from the cohort who didn't eat at the dinner, because we specifically hypothesise a food item to be the vehicle of infection in this outbreak. Excluding people reduces the sample size and therefore the power slightly, but the investigators considered that this would increase specificity.

The variables needed to define this case definition are:

-   `meal`
-   `onset_datetime`
-   `diarrhoea`
-   `bloody`
-   `vomiting`

For the case definition, we are primarily interested in three symptoms:

-   `diarrhoea` (without blood)
-   `bloody` (diarrhoea with blood)
-   `vomiting`

However, these are not the only symptoms in the data set. When creating the case definition, it would be easier to refer to these three symptoms if there was one column, indicating whether people had those symptoms or not. We can call this column `gastrosymptoms`.

Next, we created a column for the case definition, in which we defined all the respondents as either cases, non-cases or `NA`.

We can define non-cases as those who attended the meal, but didn't develop any gastro symptoms, or if they did, developed them before the meal took place.

We can exclude (define as `NA`) respondents that answered the survey, but did not attend the meal.

### Incubation times

A suitable incubation period to use in the case definition can be defined by calculating the time between exposure (to the meal) and onset of symptoms, and then looking at the distribution of these time differences. In this outbreak, incubation periods are easy to calculate, because everyone was exposed at (roughly) the same time and on the same day (eating the meal at the school dinner party).

We can see that the onset dates of all cases were from 11 to 13 November 2006 inclusive; this is why case numbers didn't change when we updated the case definition.

### Exclusions

Ultimately, the investigation team decided to remove respondents that did not meet the definition for a case or a non-case from the data set prior to analysis.
:::

# Descriptive Analyses

::: callout-important
## Instructions

In this section, participants should...
:::

::: panel-tabset
## Introduction

### Goals

In this session, we will perform some descriptive analysis. Typically, we would describe the outbreak in terms of time, place and person. In this data set, we don't have geospatial information, but we do have time (onset dates and times, incubation periods) and person (age, sex, clinical symptoms). You will:

-   Describe cases by age and sex (create an age sex pyramid)
-   Describe case symptoms
-   Describe and illustrate incubation periods
-   Create an epidemic curve
-   Calculate attack proportions
-   summarise the analysis so far and generate hypotheses:
    -   Any surprising findings regarding cohort characteristics?
    -   Does the shape of the epicurve support a viral or toxic aetiology?
    -   What other information can you obtain from the epicurve?
    -   Does the summary of symptoms help you narrow down the suspect agent?
    -   Do you have any hypothesis about the vehicle of infection so far?

Many of the descriptive features can be illustrated graphically, for which we will use the package `ggplot2`.

## Analisys Plan

In theory, you create an analysis plan before data collection - it is an important way to ensure you collect all the data you need and that you *only* collect the data you need (think: data protection!). However, during outbreak investigations, you may need to deploy questionnaires as soon as possible and before a plan has been developed. In these cases, your experience will be an important resource to fall back on.

In your analysis plan you should create a document to include:

-   research question(s) and hypothesis
-   dataset(s) to be used
-   inclusion / exclusion criteria for records you will use in your analysis
-   list of variables that will be used in the main analysis;
    -   outcome variable (being a case)
    -   main exposure variables (e.g. food consumed)
    -   stratifying variables (e.g. age, sex, group, class)
-   statistical methods
-   key mock-up tables including title, labels and expected format for results
    -   univariable
    -   bivariable
    -   stratified

## Time

## Person
:::

# References

1.  [EPIET Copenhagen outbreack case study](https://github.com/EPIET/OutbreakInvestigation/tree/master/Copenhagen)

2.  Morgenstern, H., Kleinbaum, D. G., & Kupper, L. L. (1980). [Measures of Disease Incidence Used in Epidemiologic Research. International Journal of Epidemiology, 9(1), 97--104.](https://doi.org/10.1093/ije/9.1.97)

3.  Rothman, K. J., Lash, T. L., VanderWeele, T. J., & Haneuse, S. (2021). Modern epidemiology (Fourth edition). Wolters Kluwer.

4.  Wickham, H. (n.d.). [Tidy Verse - R packages for data science.](https://www.tidyverse.org/)

5.  Wickham, H. (2009). [Ggplot2: elegant graphics for data analysis. Springer.](https://ggplot2-book.org/)

6.  Wickham, H. (2019). [Advanced R (Second edition). CRC Press/Taylor and Francis Group.](https://adv-r.hadley.nz/)

7.  Wickham, H., & Grolemund, G. (2016). [R for data science: import, tidy, transform, visualize, and model data (First edition). O'Reilly.](https://r4ds.had.co.nz/)
