# Overview -----------------

# Overview {style="color:darkblue;"}

## 

::: columns
::: {.column width="50%"}
### Questions

-   c
:::

::: {.column width="50%"}
### Lesson Objectives

#### To know the course

-   a

#### To be able to do/use

-   b
:::
:::

# topic --------------------

# Topic

## Chapter

::: callout-tip
:::

::: callout-caution
## Your turn
:::

# Code ---------------------

::: columns
::: {.column width="35%"}
|     |      |
|-----|------|
|     | \`\` |
:::

::: {.column width="65%"}
```{r}
#| output-location: column
#| results: hold

```
:::
:::

# Columns ------------------

::: columns
::: {.column width="50%"}
:::

::: {.column width="50%"}
:::
:::

# Summary ------------------

# Summary

::: columns
::: {.column width="50%"}
::: callout-important
:::
:::

::: {.column width="50%"}
::: callout-caution
## Your turn

Connect to our [pad](https://bit.ly/ubep-rws-pad)(<https://bit.ly/ubep-rws-pad>) and ...
:::
:::
:::

# Homework

## Posit's RStudio Cloud Workspace

**Instructions**

::: columns
::: {.column width="50%"}
-   Go to: https://bit.ly/ubep-rws-rstudio

::: callout-caution
## Your turn

-   Project: day-1
-   Instructions:
    -   Go to: https://bit.ly/ubep-rws-website
    -   The text is the Day-2 assessment under the tab "Summative Assessments".
    -   (on RStudio Cloud) `homework/day_one-summative.html`
-   Script to complete: `homework/solution.R`
:::
:::
:::

# Break --------------------

## Break

[20 minutes](https://apps.garrickadenbuie.com/countdown/?_inputs_&time=%2220%3A00%22&update_every=%221%22&warn_time=%2204%3A00%22)[^templates-1]

[^templates-1]: <https://apps.garrickadenbuie.com/countdown>

# Your turn ----------------

## My turn

YOU: Connect to our [pad](https://bit.ly/ubep-rws-pad) (<https://bit.ly/ubep-rws-pad>) and write there questions & doubts (and if I am too slow or too fast)

ME: Connect to the Course-script project in RStudio cloud (<https://bit.ly/ubep-rws-rstudio>): script `17-gtsummary.R`

## Your turn

::: callout-caution
## Your turn

-   Connect to our [pad](https://bit.ly/ubep-rws-pad)(<https://bit.ly/ubep-rws-pad>)

-   Connect to the Day-4 project in RStudio cloud (<https://bit.ly/ubep-rws-rstudio>)

...and:
:::

1.  Before to evaluate it, in the pad, under the section `4.2. Ex29`, write (in a new line) what is your expected result from the following computation:

``` r
<code>
```

2.  Then, open the script `19-tbl_cross.R` and follow the instruction step by step.

::: {.fragment .fade-out fragment-index="1"}
```{r}
#| echo: false
countdown(
  minutes = 3,
  left = "25%",
  right = "25%",
  top = "70%",
  bottom = 0,
  blink_colon = TRUE,
  play_sound = TRUE,
  margin = "5%",
  font_size = "4em"
)
```
:::

::: {.fragment .fade-in fragment-index="1"}
::: callout-important
-   `a:b` create a sequence of from `a` to `b`
-   `a` and `b` can be zero or negative
-   The sequence start from `a` and go through `b` in by steps of length `1`, without surpasses `b`
-   Math and logic works on vectors (i.e., sequences) input and output!
:::
:::
